---
title: "Bayesian Profile Regression"
author: "Scott Pease"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(PReMiuM)
```

# Load Data
```{r}
all_data = readr::read_csv("/Users/Scott/QGIS/SVI/Penn_Analysis/data/final_output.csv")
```

# Remove Wonky Rows
remove that Lake Erie (the literal lake) tract:
```{r}
all_data <- all_data |> filter(!is.na(COUNTY))
```

## remove empty (Zero pop) tracts
Removing TL_sum_pop_est zeroes gets rid of NAs in the outcome
Removing E_TOTPOP zeroes: the covariates are NA for these rows
b/c ACS says nobody's there:
`x = all_data |> 
  filter(E_TOTPOP == 0) |> 
  select(all_of(covnames)
`
SO I'll remove BOTH.
```{r}
all_data <- all_data |> filter(E_TOTPOP > 0)
all_data <- all_data |> filter(TL_sum_pop_est > 0)
```


# Prep Columns
## Outcome
```{r}
all_data = all_data |> 
  mutate(
    WWR_sum_pop_est = tidyr::replace_na(WWR_sum_pop_est, 0),
    frac_WW_CEISIN = WWR_sum_pop_est / TL_sum_pop_est
  )
```

```{r}
all_data = all_data |> 
  mutate(
    WW_pop_OUTCOME = E_TOTPOP * frac_WW_CEISIN
  )
```

## NEW CHANGE:
removing fully PWS-based tracts
reason: BPR was crap
```{r}
all_data <- all_data |> filter(WW_pop_OUTCOME > 0)
```


## Rurality Confounder
Should pop density be based on land area only, or total area (i.e. including water area as well)?
```{r}
all_data = all_data |> 
  mutate(
    pop_density = E_TOTPOP / ALAND,
    pop_density_alt = E_TOTPOP / (ALAND + AWATER),
    )
```

## Clustering Covariates
```{r}
covnames = c(
    "EP_POV150", 
    "EP_UNEMP",
    "EP_NOHSDP",
    "EP_AGE65",
    "EP_AGE17",
    "EP_MINRTY",
    "EP_HBURD"
)
```

Set NA's (are encoded -999) for covariates:
ref: https://stackoverflow.com/questions/42888008/changing-multiple-column-values-given-a-condition-in-dplyr
ref: https://stackoverflow.com/questions/67643144/where-is-the-purrr-operator-documented
```{r}
all_data = all_data |> 
  mutate(
    across(
      all_of(covnames),
      ~ if_else(.x == -999, NA, .x)
    )
  )
```

Convert to quartiles
```{r}
for (nom in covnames) {
  qs = quantile(all_data[nom], na.rm = TRUE)
  all_data = all_data |>
    rowwise() |> 
    mutate("{nom}" := if_else(.data[[nom]] == qs[1], 1, sum(.data[[nom]] > qs))) |> 
    ungroup()
}
```

# Select Data
ID = GEOID,
outcome = WW_pop_OUTCOME,
offset (outcomeT) = E_TOTPOP
confounders = pop_density,
clustering covariates = covnames
```{r}
input_data = all_data |> 
  select(GEOID, E_TOTPOP, WW_pop_OUTCOME, pop_density, all_of(covnames))
```


# BPR
```{r}
mod_no_zero_WW = PReMiuM::profRegr(
  covNames = covnames,
  fixedEffectsNames = c("pop_density"),
  outcome = "WW_pop_OUTCOME",
  outcomeT = "E_TOTPOP",
  data = input_data,
  output = "bpr_output/mod_no_zero_WW",
  nClusInit = 15,  # c.f. Wamsley
  yModel = "Poisson",
  xModel = "Discrete",
  extraYVar = TRUE,  # for NegBinomial
  # includeCAR = TRUE
)
```

```{r}
dissimmat_mod_no_zero_WW = calcDissimilarityMatrix(mod_no_zero_WW)
```

```{r}
optimal_clust_mod_no_zero_WW = calcOptimalClustering(dissimmat_mod_no_zero_WW)
```

```{r}
# input_data["cluster_mod_no_zero_WW"] = as.factor(optimal_clust_mod_no_zero_WW[["clustering"]])
```

```{r}
riskProfileObj_mod_no_zero_WW <- calcAvgRiskAndProfile(optimal_clust_mod_no_zero_WW)
```

```{r}
clusterOrderObj_mod_no_zero_WW <- plotRiskProfile(riskProfileObj_mod_no_zero_WW, "risk_profile_mod_no_zero_WW.png")
```

