---
title: "BPR_CAR"
author: "Scott Pease"
date: "2025-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(PReMiuM)
library(sf)
library(spdep)
```

# SVI Data
Need to figure out which cols to remove from the neighbors matrix stuff 
with {spdep}
## Load Data
```{r}
all_data = readr::read_csv(
  "/Users/Scott/QGIS/SVI/Penn_Analysis/data/final_output.csv",
  col_types = list(
    GEOID = col_character()
  ))
```

# Remove Wonky Rows
remove that Lake Erie (the literal lake) tract:
```{r}
all_data <- all_data |> filter(!is.na(COUNTY))
```

## remove empty (Zero pop) tracts
Removing TL_sum_pop_est zeroes gets rid of NAs in the outcome
Removing E_TOTPOP zeroes: the covariates are NA for these rows
b/c ACS says nobody's there:
`x = all_data |> 
  filter(E_TOTPOP == 0) |> 
  select(all_of(covnames)
`
SO I'll remove BOTH.
```{r}
all_data <- all_data |> filter(E_TOTPOP > 0)
all_data <- all_data |> filter(TL_sum_pop_est > 0)
```

## VERY IMPORTANT
THIS LINES UP WITH THE INLA FILE, PROVIDED THAT IS ALSO SORTED BY GEOID
```{r}
all_data <- all_data |> arrange(GEOID)
```


# TIGER/Line
```{r}
TL_data = st_read("/Users/Scott/QGIS/SVI/Penn_Analysis/data/tl_2022_42_tract/tl_2022_42_tract.shp")
```

```{r}
class(TL_data)
```

```{r}
non_zero_pop_tracts = all_data$GEOID
TL_data = TL_data |> 
  filter(GEOID %in% non_zero_pop_tracts) |> 
  arrange(GEOID)
```


# Neighbors Stuff
```{r}
TL_nb = poly2nb(TL_data, row.names = TL_data$GEOID)
```

```{r}
coords <- st_coordinates(st_centroid(st_geometry(TL_data)))
```


```{r}
plot(st_geometry(TL_data))
plot(TL_nb, coords, add=TRUE, col="red")
```

```{r}
nb2INLA("INLA_file", TL_nb)
```
