---
title: "corrs"
author: "Scott Pease"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Load Data
```{r}
all_data = readr::read_csv("/Users/Scott/QGIS/SVI/Penn_Analysis/data/final_output.csv")
```

# Remove Wonky Rows
remove that Lake Erie (the literal lake) tract:
```{r}
all_data <- all_data |> filter(!is.na(COUNTY))
```

## remove empty (Zero pop) tracts
Removing TL_sum_pop_est zeroes gets rid of NAs in the outcome
Removing E_TOTPOP zeroes: the covariates are NA for these rows
b/c ACS says nobody's there:
`x = all_data |> 
  filter(E_TOTPOP == 0) |> 
  select(all_of(covnames)
`
SO I'll remove BOTH.
```{r}
all_data <- all_data |> filter(E_TOTPOP > 0)
all_data <- all_data |> filter(TL_sum_pop_est > 0)
```


# Prep Columns
## Outcome
```{r}
all_data = all_data |> 
  mutate(
    WWR_sum_pop_est = tidyr::replace_na(WWR_sum_pop_est, 0),
    frac_WW_CEISIN = WWR_sum_pop_est / TL_sum_pop_est
  )
```

```{r}
all_data = all_data |> 
  mutate(
    WW_pop_OUTCOME = E_TOTPOP * frac_WW_CEISIN
  )
```

## Rurality Confounder
Should pop density be based on land area only, or total area (i.e. including water area as well)?
```{r}
all_data = all_data |> 
  mutate(
    pop_density = E_TOTPOP / ALAND,
    pop_density_alt = E_TOTPOP / (ALAND + AWATER),
    )
```

## Clustering Covariates
```{r}
covnames = c(
    "EP_POV150", 
    "EP_UNEMP",
    "EP_NOHSDP",
    "EP_AGE65",
    "EP_AGE17",
    "EP_MINRTY",
    "EP_HBURD"
)
```

Set NA's (are encoded -999) for covariates:
ref: https://stackoverflow.com/questions/42888008/changing-multiple-column-values-given-a-condition-in-dplyr
ref: https://stackoverflow.com/questions/67643144/where-is-the-purrr-operator-documented
```{r}
all_data = all_data |> 
  mutate(
    across(
      all_of(covnames),
      ~ if_else(.x == -999, NA, .x)
    )
  )
```

# County-level z-scoring
```{r}
all_data = all_data |> 
  mutate(
    across(
      all_of(covnames),
      ~ as.vector(scale(.x)),
      .names = "{.col}_allPA_std"
    )
  ) |> 
  group_by(COUNTY) |> 
  mutate(
    across(
      all_of(covnames),
      ~ as.vector(scale(.x)),
      .names = "{.col}_county_std"
    )
  ) |> 
  ungroup()
```



Convert to quartiles
```{r}
for (nom in covnames) {
  for (suffix in c("", "_allPA_std", "_county_std")) {
    full_nom = paste(nom, suffix, sep = "")
    qs = quantile(all_data[full_nom], na.rm = TRUE)
    all_data = all_data |>
      rowwise() |> 
      mutate("{full_nom}_quartile" := if_else(.data[[full_nom]] == qs[1], 1, sum(.data[[full_nom]] > qs))) |> 
      ungroup()
  }
}
```

```{r}
all.equal(all_data$EP_POV150_quartile, all_data$EP_POV150_allPA_std_quartile)
```


```{r}
covnames_allPA = paste0(covnames, "_allPA_std")
covnames_county = paste0(covnames, "_county_std")
```

# Select Data
ID = GEOID,
outcome = WW_pop_OUTCOME,
offset (outcomeT) = E_TOTPOP
confounders = pop_density,
clustering covariates = covnames
```{r}
input_data = all_data |> 
  select(WW_pop_OUTCOME, frac_WW_CEISIN, pop_density, all_of(covnames), all_of(covnames_allPA), all_of(covnames_county))
```

# Plots
```{r}
cor(tidyr::drop_na(input_data))
```

```{r}
for (nom in covnames_county) {
  show(input_data |> 
    ggplot() +
    geom_point(aes(.data[[nom]], frac_WW_CEISIN)))
}
```

```{r}
input_data["x"] = exp(input_data["EP_POV150"] / 100)
```

Does putting in on the outcome scale help at all?
By which I mean, E(y) = lambda, log(lambda) = Xb, so maybe try exp(X)
```{r}
for (nom in covnames) {
  show(input_data |> 
    ggplot() +
    geom_point(aes(exp(.data[[nom]] / 100), frac_WW_CEISIN)))
}
```

```{r}
for (nom in covnames) {
  x = input_data |> filter(WW_pop_OUTCOME > 20)
  show(x |> 
    ggplot() +
    geom_point(aes(WW_pop_OUTCOME, .data[[nom]])))
}
```



```{r}
for (nom in covnames) {
  x = input_data |> filter(frac_WW_CEISIN > 0.025, frac_WW_CEISIN < 0.975)
  show(x |> 
    ggplot() +
    geom_point(aes(frac_WW_CEISIN, .data[[nom]])))
}
```


```{r}
input_data |> 
    ggplot() +
    geom_point(aes(log(pop_density), frac_WW_CEISIN))
```


