---
title: "corrs"
author: "Scott Pease"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(ggbiplot)
```

# Load Data
## County Rural/Urban Data
```{r}
county_type_data = readr::read_csv(
  "../data/List of Counties web 2022.csv", 
  col_types = cols(
    Rural_Urban = col_factor(levels = c("Rural", "Urban"))
  )
)
```

## remaining data
```{r}
all_data = readr::read_csv("/Users/Scott/QGIS/SVI/Penn_Analysis/data/final_output.csv")
```

# Remove Wonky Rows
remove that Lake Erie (the literal lake) tract:
```{r}
all_data <- all_data |> filter(!is.na(COUNTY))
```

## remove empty (Zero pop) tracts
Removing TL_sum_pop_est zeroes gets rid of NAs in the outcome
Removing E_TOTPOP zeroes: the covariates are NA for these rows
b/c ACS says nobody's there:
`x = all_data |> 
  filter(E_TOTPOP == 0) |> 
  select(all_of(covnames)
`
SO I'll remove BOTH.
```{r}
all_data <- all_data |> filter(E_TOTPOP > 0)
all_data <- all_data |> filter(TL_sum_pop_est > 0)
```


# Prep Columns
## Outcome
```{r}
all_data = all_data |> 
  mutate(
    WWR_sum_pop_est = tidyr::replace_na(WWR_sum_pop_est, 0),
    frac_WW_CEISIN = WWR_sum_pop_est / TL_sum_pop_est
  )
```

```{r}
all_data = all_data |> 
  mutate(
    WW_pop_OUTCOME = E_TOTPOP * frac_WW_CEISIN
  )
```


## County Confounder
```{r}
all_data = all_data |> full_join(
  county_type_data,
  by = join_by(COUNTY == County_Full_Name),
  relationship = "many-to-one"
)
```

## Rurality Confounder
```{r}
all_data = all_data |> 
  mutate(
    pop_density = E_TOTPOP / ALAND,
    )
```

## Clustering Covariates
```{r}
covnames = c(
    "EP_POV150", 
    "EP_UNEMP",
    "EP_NOHSDP",
    "EP_AGE65",
    "EP_AGE17",
    "EP_MINRTY",
    "EP_HBURD"
)
```

Set NA's (are encoded -999) for covariates:
ref: https://stackoverflow.com/questions/42888008/changing-multiple-column-values-given-a-condition-in-dplyr
ref: https://stackoverflow.com/questions/67643144/where-is-the-purrr-operator-documented
```{r}
all_data = all_data |> 
  mutate(
    across(
      all_of(covnames),
      ~ if_else(.x == -999, NA, .x)
    )
  )
```

# Select Data
ID = GEOID,
outcome = WW_pop_OUTCOME,
offset (outcomeT) = E_TOTPOP
confounders = pop_density,
clustering covariates = covnames
```{r}
input_data = all_data |> 
  select(COUNTY, Rural_Urban, frac_WW_CEISIN, pop_density, all_of(covnames)) |> 
  tidyr::drop_na()
```

# prcomp
```{r}
mod = prcomp(input_data[,5:9], scale. = TRUE)
```

```{r}
autoplot(mod, data = input_data, color = "frac_WW_CEISIN", loadings = TRUE, loadings.label = TRUE)
```

```{r}
mod_w_popdens = prcomp(input_data[,4:9], scale. = TRUE)
```

```{r}
autoplot(mod_w_popdens, data = input_data, color = "frac_WW_CEISIN", loadings = TRUE, loadings.label = TRUE, )
```

```{r}
biplot(mod, scale = 0)
biplot(mod, scale = 1)
biplot(mod, scale = 1, pc.biplot = TRUE)
biplot(mod, scale = 1, pc.biplot = TRUE, )
```

```{r}
ggbiplot::ggbiplot(mod, varname.color = "red", groups = input_data$Rural_Urban, ellipse = TRUE, point.size = 0.25)
ggbiplot::ggbiplot(mod, varname.color = "red", groups = input_data$frac_WW_CEISIN, point.size = 0.25, alpha = 0.75) + scale_colour_gradient(low = "yellow", high = "red") + guides(color=guide_colorbar(title="Prop Well Water Users"))
```

